---
title: "Getting Started with R Markdown"
subtitle: "Beginner to intermediate level tutorial for R Markdown"
author: "Dave Bosworth<br>EMRR Branch<br>Collaborative Science and Innovation Section"
date: "October 17, 2022"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: [default, metropolis, metropolis-fonts, "css/custom.css"]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
---

```{r setup, include = FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

# Introduction

- A little about me

--

- About you

---

# Tutorial Outline

--

R Markdown basics: What is it, and why should I use it?  
**Activity 1:** Open test file, load data, and run code  

--

Basic syntax and structure  
**Activity 2:** Create new document, practice using some syntax options, knit file to html 

--

Code chunks and inline code  
**Activity 3:** Inserting and running code  

--

Code chunk options  
**Activity 4:** Modify code chunk options  

--

Basic Pandoc-header options  
**Activity 5:** Modify YAML options  

---

# Tutorial Outline - continued

--

Adding and formatting tables  
**Activity 6:** Tables  

--

Adding interactive elements: plotly, leaflet, datatable (DT)  
**Activity 7:** Interactive elements  

--

Adding tabs  
**Activity 8:** Tabs  

--

Open Questions  

---

# What is R Markdown?

- R-specific file type (.Rmd) that provides a platform for science communication

--

- Used to save and execute code just like an R script, but in a notebook format

--

- Allows for code and text narrative in the same document

---

# How is R Markdown different than an R script?

- Results from code and narrative text can be stitched ("knitted") together in the output

--

- Allows for many output options including: 
  - html, pdf, and Microsoft Word 
  - presentation slides
  - books and scientific articles
  - dashboards and shiny applications
  - websites

---

# Why use R Markdown?

- Organize your code into "chunks" that can be run individually (notebook format)

--

- Save code and narrative text together in one document

--

- Share code and results with others - reproducible, transparent, clearly communicated

--

- Can be used to create automated reports

---
class: inverse, middle

# Activity 1

Open test.Rmd file, load data, and run code

---

# R Markdown Document Structure

## Metadata (YAML header)

Optional section for rendering options (pandoc) as key: value pairs

```{r example yaml header, eval = FALSE}
---
title: "Getting Started with R Markdown"
subtitle: "Beginner to intermediate level tutorial for R Markdown"
author: "Dave Bosworth<br>EMRR Branch<br>Collaborative Science and Innovation Section"
date: "October 17, 2022"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: [default, metropolis, metropolis-fonts, "css/custom.css"]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
---
```

---

# R Markdown Document Structure

## Narrative Text

Narration formatted with markdown-

- Plain text
- Headers
- Text formatting (bold, italic)
- Links
- Images
- Numbered and unordered lists
- Code
- Equations

[R Markdown Reference Guide](https://www.rstudio.com/wp-content/uploads/2015/03/rmarkdown-reference.pdf)

---

# R Markdown Document Structure

## Code chunks

- Sections of embedded R code within the document
- Code is run when R Markdown file is "knitted" with results displayed in the output document 

```{r code chunk example, fig.height = 5}
plot(pressure)
```

---
class: inverse, middle

# Activity 2

- Create new document
- Practice using some syntax options
- "Knit" file to html

---

# Code within R Markdown

### Code chunks

```{r code chunk example 2}
summary(cars)
```

--

***

### Inline code

```{r inline code example}
# Today's date is `r Sys.Date()`
```

Today's date is `r Sys.Date()`

---
class: inverse, middle

# Activity 3

Inserting and running code

---

# Code Chunk Options

### message, error, warning

message = TRUE

```{r true message output example, message = TRUE}
library(tidyverse)
```

--

***

message = FALSE

```{r false message output example}
library(tidyverse)
```

---

# Code Chunk Options

### eval, include, echo, results

- eval: if **FALSE**, the code will not run in the code block when "knitted", but the code appears in the output file (default is **TRUE**)

--

- include: if **FALSE**, the code and results won't appear in output file, but the code is still executed and results can be used by other code chunks (default is **TRUE**)

--

- echo: if **FALSE**, the code won't appear in the output file, but the results are still displayed (default is **TRUE**)

--

- results: if **"hide"**, the results won't appear in the output file, but the code appears in the output (default is **"markup"**)

---

# Code Chunk Options

### Figure output options

- fig.height, fig.width: the height and width in inches for figures created by the code chunk

--

- fig.align: allows for custom alignment of figures in the output file. Can be "left", "right", or "center"

---
class: inverse, middle

# Activity 4

Modify code chunk options

---

# Basic Pandoc-header options

---
class: inverse, middle

# Activity 5

Modify YAML options

---
