---
title: "R Markdown Tutorial - Activity Examples"
author: "Dave Bosworth"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
  html_document:
    code_folding: show
    code_download: true
    toc: true
    toc_float:
      collapsed: false
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Purpose

Introduce and display examples of how to use R Markdown to bring together code and text in one document to share your work.

# Global code and functions

```{r load packages}
# Load packages
library(tidyverse)
library(knitr)
library(here)
```

# Import Data

Let's import some example fish count data collected by YBFMP. 

```{r import data}
df_monthly <- read_csv(here("materials/Monthly_Catch_CPUE.csv"))
```

This data contains monthly fish counts and CPUE for three different gear types (`r unique(df_monthly$Gear)`) from `r min(df_monthly$year)` to `r max(df_monthly$year)`.

***

**A word about working directories and .Rmd files:**  
The default working directory when an .Rmd file is knitted is the .Rmd document directory. This can get confusing when you are using an R project, in which the default working directory is the root directory of the R project. So it's possible to have different working directories when you are running code from code blocks in an .Rmd file versus when you knit the .Rmd file. A useful R package to help with working directories and file paths is the [`here` R package](https://here.r-lib.org/). You can see that we're using it in the code chunk above to clearly define the file path for the fish data.  The [`here::here()`](https://here.r-lib.org/reference/here.html) function points to the root directory of an R project regardless of what the current working directory is. For example:

```{r here function example}
# Current working directory
getwd()

# Working directory defined by here::here()
here()
```

You can continue the file path within the `here::here()` function as a character string, which appends this string to the the root directory of the R project. For example, here is the file path to the fish data imported above:

```{r here function example 2}
# Define file path to fish data relative to the root directory using here::here()
here("materials/Monthly_Catch_CPUE.csv")
```

# Using Code Chunks

Let's practice using code chunks to create and print a plot displaying monthly CPUE by gear type.

```{r cpue plot}
plt_cpue_month <- df_monthly %>% 
  ggplot(aes(x = Date, y = CPUE)) +
  geom_col() +
  facet_wrap(vars(Gear), ncol = 1, scales = "free_y") + 
  theme_bw()

plt_cpue_month
```

# Code Chunk Options

Now that we've seen how to use code chunks in an .Rmd document, lets take a look at some commonly-used code chunk options that configure if and how the code chunk and its results are displayed.

## message, warning

Notice when we imported the fish count data above, there is a message printed in the output. If we don't want this message to display, we can set `message = FALSE` for the code chunk.

```{r import data 2, message = FALSE}
df_monthly <- read_csv(here("materials/Monthly_Catch_CPUE.csv"))
```

Now the message doesn't display, but the code was still executed. Same applies to warnings.

## eval

If you want the code to appear in the output file, but for the **code to NOT run** when the Rmd file is knitted, set `eval = FALSE` as a code chunk option. For example, you may want to show the code you used to summarize the count and CPUE data as annual totals in a new data frame `df_annual` and plot the CPUE results saving it as `plt_cpue_yr`, but not actually run the code when you knit the file.

```{r eval false example, eval = FALSE}
# Summarize count and CPUE as annual totals
df_annual <- df_monthly %>% 
  group_by(year, Gear) %>% 
  summarize(across(c(mo.count, CPUE), sum), .groups = "drop") %>%
  rename(
    Year = year,
    Count = mo.count
  )

# Plot annual CPUE
plt_cpue_yr <- df_annual %>% 
  ggplot(aes(x = Year, y = CPUE)) +
  geom_col() +
  facet_wrap(vars(Gear), ncol = 1, scales = "free_y") + 
  theme_bw()

plt_cpue_yr
```

Now if you were to try to print `df_annual` and `plt_cpue_yr`, you'll get an error since the code wasn't executed when the document was knitted.

```{r error df_annual and plt_cpue_yr, error = TRUE}
# Show glimpse of df_annual
glimpse(df_annual)

# Print plot of annual CPUE
plt_cpue_yr
```

## include

If you want the code to be executed when the Rmd file is knitted, but for the **code and results to NOT appear** in the output file, set `include = FALSE` as a code chunk option. For example, let's summarize and plot the CPUE data using the same code shown above, but this time we'll use `include = FALSE` so the code and results won't appear in the output document.

```{r include false example, include = FALSE}
# Summarize count and CPUE as annual totals
df_annual <- df_monthly %>% 
  group_by(year, Gear) %>% 
  summarize(across(c(mo.count, CPUE), sum), .groups = "drop") %>%
  rename(
    Year = year,
    Count = mo.count
  )

# Plot annual CPUE
plt_cpue_yr <- df_annual %>% 
  ggplot(aes(x = Year, y = CPUE)) +
  geom_col() +
  facet_wrap(vars(Gear), ncol = 1, scales = "free_y") + 
  theme_bw()

plt_cpue_yr
```

Even though the code and results aren't displayed in the document, the `df_annual` data frame and `plt_cpue_yr` plot object are available to be used within the Rmd file.

```{r show df_annual and plt_cpue_yr}
# Show glimpse of df_annual
glimpse(df_annual)

# Print plot of annual CPUE
plt_cpue_yr
```

## echo

If you want the code to be executed and results to be displayed in the output file when the Rmd is knitted, but for the **code to NOT appear** in the output file, set `echo = FALSE` as a code chunk option. For example, maybe you want `glimpse(df_annual)` and the plot of annual CPUE to display, but not show the code used to execute this.

```{r echo false example, echo = FALSE}
# Show glimpse of df_annual
glimpse(df_annual)

# Print plot of annual CPUE
plt_cpue_yr
```

## Figure output options

The default figure dimensions in a rendered R Markdown document is 7 inches wide and 5 inches tall. Sometimes you'll want to change these, which can be done using the `fig.height` and `fig.width` code chunk options. Let's change the dimensions of the annual CPUE figure to 6 by 6.

```{r plot annual cpue change dims, fig.height = 6, fig.width = 6}
plt_cpue_yr
```

Now, maybe we want the figure to be aligned on the right-side of the knitted document. In this case, we'll use the `fig.align` code chunk option.

```{r plot annual cpue change align, fig.height = 6, fig.width = 6, fig.align = 'right'}
plt_cpue_yr
```

If we want to add a caption to our figure, we'll use the `fig.cap` code chunk option.

```{r plot annual cpue add caption, fig.cap = "Figure 1: This is our annual CPUE figure."}
plt_cpue_yr
```

Finally, if we want to add alt text to our figure, we'll use the `fig.alt` option.

```{r annual cpue alt text, include = FALSE}
plt_cpue_yr_altext <- "This figure shows the annual CPUE for each gear type used by the YBFMP from 1998-2020."
```

```{r plot annual cpue add alt text, fig.alt= plt_cpue_yr_altext}
plt_cpue_yr
```

# Tables

This is what it looks like when you print a data frame itself.

```{r annual data as tibble}
df_annual
```

`knitr::kable()` offers a simple format for displaying tables in a rendered R Markdown document. For example, here is the table of annual count and CPUE data printed as a kable.

```{r annual data as kable}
kable(df_annual, caption = "Annual Counts and CPUE of YBFMP survey data")
```

This is pretty nice compared to just printing the data frame itself. However, the kable formatting can be improved. The `knitr::kable()` function allows for some minor formatting such as rounding digits and column alignment.

```{r annual data as kable with format}
kable(
  df_annual, 
  caption = "Annual Counts and CPUE of YBFMP survey data",
  # Round CPUE column to 1 digit
  digits = 1,
  # Align all columns to the left
  align = "l"
)
```

The `kableExtra` R package allows for many more formatting options for kables. As of version 1.2+, the author of `kableExtra` recommends using the `kbl()` function to create the kable.

```{r format kable}
library(kableExtra)

df_annual %>%
  kbl(
    caption = "Annual Counts and CPUE of YBFMP survey data",
    # Round CPUE column to 1 digit
    digits = 1,
    # Align all columns to the left
    align = "l"
  ) %>% 
  kable_styling(
    # appearance option
    bootstrap_options = c("striped", "hover"),
    # change width
    full_width = FALSE,
    # change position
    position = "left",
    # fix the header row at the top when scrolling, useful for longer tables
    fixed_thead = TRUE
  ) %>%
  # Format the header row (0) to be aligned in the center
  row_spec(0, align = "center") %>%
  # Make the Year column bold
  column_spec(1, bold = TRUE) %>%
  # add a scroll box for easier viewing
  scroll_box(width = "40%", height = "550px")
```
<br>

You can do so much more with the `kableExtra` R package. Check out the package  [documentation](https://haozhu233.github.io/kableExtra/) for more details and options.

# Interactive Elements

There are a whole lot of R packages that allow for some really nice interactive widgets in R Markdown. We'll cover three of these that I've used and found helpful.

## datatable

You can convert a data frame to an interactive table using the `DT::datatable()` function.

```{r using datatable}
library(DT)
datatable(df_annual)
```

There are also some options that you can set within the `DT::datatable()` function.

```{r datatable with options}
df_annual %>% 
  # Convert Year and Gear variables to factors to allow for select lists in the
    # datatable filters
  mutate(across(c(Year, Gear), factor)) %>% 
  datatable(
    # remove rownames
    rownames = FALSE,
    # add column filters
    filter = "top",
    options = list(autoWidth = TRUE)
  )
```

## plotly

You can create interactive plots using the `plotly` R package. The `plotly::ggplotly()` function converts `ggplot2` objects into interactive plots. I found that it works best with less complicated plots.

```{r using plotly, warning = FALSE}
library(plotly)
ggplotly(plt_cpue_yr)
```

## leaflet

You can make interactive maps using the `leaflet` R package. First we'll download the fish monitoring stations for YBFMP from the [EDI data package](https://portal.edirepository.org/nis/mapbrowse?packageid=edi.233.3).

```{r download and prepare fish station data}
df_stations <- read_csv("https://portal.edirepository.org/nis/dataviewer?packageid=edi.233.3&entityid=89146f1382d7dfa3bbf3e4b1554eb5cc")

# Convert to sf object
library(sf)
sf_stations <- df_stations %>% st_as_sf(coords = c("Longitude", "Latitude"), crs = 4326)
```

Next, we'll create an interactive map of the sampling locations.

```{r using leaflet}
library(leaflet)
sf_stations %>% 
  leaflet() %>% 
  addTiles() %>%
  addCircleMarkers(
    radius = 5,
    fillOpacity = 0.8,
    weight = 0.5,
    color = "black",
    opacity = 1
  )
```

Now, we'll add labels for each station and color code the markers by the `MethodCode` (`r unique(sf_stations$MethodCode)`).

```{r leaflet with more options}
# Define color palette for MethodCode
color_pal <- colorFactor(palette = "viridis", domain = sf_stations$MethodCode)

sf_stations %>% 
  leaflet() %>% 
  addTiles() %>%
  addCircleMarkers(
    radius = 5,
    fillColor = ~color_pal(MethodCode),
    fillOpacity = 0.8,
    weight = 0.5,
    color = "black",
    opacity = 1,
    label = paste0("Station Code: ", sf_stations$StationCode)
  ) %>% 
  addLegend(
    position = "topright",
    pal = color_pal,
    values = sf_stations$MethodCode,
    title = "Gear Type:"
  )
```

# Adding Tabs

A feature that I've found to be useful is using tabs in a rendered R Markdown document. Here is how you would set up the document to display the two CPUE plots and the stations map in tabs for easier navigation. Apply the `{.tabset}` class attribute to the `## YBFMP Data Summary` header below like this: `## YBFMP Data Summary {.tabset}`. Then, each of the sub-headers of this header will appear as tabs instead of standalone sections. Use another header of the same level as `## YBFMP Data Summary` (level 2) to stop using tab navigation.

## YBFMP Data Summary {.tabset}

### Monthly CPUE

```{r monthly cpue plot tab, echo = FALSE}
plt_cpue_month
```

### Annual CPUE

```{r annual cpue plot tab, echo = FALSE}
plt_cpue_yr
```

### Stations Map

```{r stations map tab, echo = FALSE}
sf_stations %>% 
  leaflet() %>% 
  addTiles() %>%
  addCircleMarkers(
    radius = 5,
    fillColor = ~color_pal(MethodCode),
    fillOpacity = 0.8,
    weight = 0.5,
    color = "black",
    opacity = 1,
    label = paste0("Station Code: ", sf_stations$StationCode)
  ) %>% 
  addLegend(
    position = "topright",
    pal = color_pal,
    values = sf_stations$MethodCode,
    title = "Gear Type:"
  )
```

## Tab formatting

Add `.tabset-fade` and/or `.tabset-pills` to the `{.tabset}` class attribute to change the appearance and behavior of the tabs.

